# Exercises

These are intended to be done **after** completing the worked examples.



```{r}
suppressPackageStartupMessages({
  library(readr)
  library(biomaRt)
  library(dplyr)
})
```


## Exercise 1 — https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE119732

Using **GSE119732**, confirm whether the ID column contains Ensembl IDs with version suffixes.

1. Extract the first 20 IDs.
2. Count how many contain a `.`.
3. Create a new column with versions stripped.
4. Map the identifiers to HGNC symbols.

```{r}

{r, message=FALSE, warning=FALSE}


# helper function
strip_ensembl_version <- function(x) sub("\\..*$", "", x)

source("./fetch_geo_supp.R")

params_119732 <- list(
  gse       = "GSE119732",
  file_grep = "count_table_RNA_seq"
)

fetch_geo_supp(gse = params_119732$gse)

path_119732 <- file.path("data", params_119732$gse)
files_119732 <- list.files(
  path       = path_119732,
  pattern    = params_119732$file_grep,
  full.names = TRUE,
  recursive  = TRUE
)

files_119732  # should show GSE119732_count_table_RNA_seq.txt.gz

# read the count table
x1 <- read_tsv(files_119732[1], show_col_types = FALSE)

# the ID column is the first column
id_col1 <- names(x1)[1]
ids1    <- as.character(x1[[id_col1]])

## 1. Extract the first 20 IDs.
first20_1 <- ids1[1:20]
first20_1

## 2. Count how many contain a "."
sum(grepl("\\.", first20_1))

## 3. Create a new column with versions stripped.
x1$ID_stripped <- strip_ensembl_version(ids1)

head(x1[, c(id_col1, "ID_stripped")], 10)

## 4. Map the identifiers to HGNC symbols.

ensembl_ids1 <- unique(x1$ID_stripped)

ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
map1 <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = ensembl_ids1,
  mart       = ensembl
)

expr_mapped_1 <- x1 %>%
  mutate(ensembl_gene_id = ID_stripped) %>%
  left_join(
    map1,
    by = "ensembl_gene_id",
    relationship = "many-to-many"  # acknowledge non 1:1 mapping
  )

head(expr_mapped_1[, 1:8])
```


## Exercise 2 — https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122380

Using **GSE122380**, confirm whether the ID column contains Ensembl IDs with version suffixes.

1. Extract the first 20 IDs.
2. Create a new column with versions stripped.
3. Map the identifiers to HGNC symbols.
4. What is different about this file? 


```{r}

params_122380 <- list(
  gse       = "GSE122380",
  file_grep = "raw_counts"
)

fetch_geo_supp(gse = params_122380$gse)

path_122380 <- file.path("data", params_122380$gse)

files_122380 <- list.files(
  path       = path_122380,
  pattern    = params_122380$file_grep,
  full.names = TRUE,
  recursive  = TRUE
)

files_122380  # should be "data/GSE122380/GSE122380_raw_counts.txt.gz"

# read the raw counts

x2 <- read_table(
  files_122380[1],
  comment = "#",    
  col_types = cols()  
)

# test
#dim(x2)
#head(x2[, 1:5])
#names(x2)

## 1) Extract the first 20 IDs

id_col2 <- names(x2)[1]      # gene ID column
ids2    <- as.character(x2[[id_col2]])

first20_2 <- ids2[1:20]
first20_2

## 2) Create a new column with versions stripped.

x2$ID_stripped <- strip_ensembl_version(ids2)
head(x2[, c(id_col2, "ID_stripped")], 10)

## 3) Map the identifiers to HGNC symbols.

ensembl_ids2 <- unique(x2$ID_stripped)

ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
map2 <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = ensembl_ids2,
  mart       = ensembl
)

str(map2$ensembl_gene_id)  

expr_mapped_2 <- x2 %>%
  mutate(ensembl_gene_id = ID_stripped) %>%
  left_join(map2, by = "ensembl_gene_id")  

head(expr_mapped_2[, 1:8])
head(expr_mapped_2[, c("Gene_id", "ID_stripped", "hgnc_symbol")], 20)

```
4. what is different about this file is that the version numbers are already stripped so my stripped helper function does nothing, none of the ids contain a dot.
For GSE122380, the raw counts file is whitespace delimited, so i had to change the way i read the table using read table instead of read tsv which is the first way i tried.



## Exercise 3 - 
 
Can you use the worked example to process the above two GEO records?  How?

i used the worked example to practice processing the geo records. i was inspired by the workflow and i used the same steps. i downloaded the supplementary data. i used read tsv for the first exercise, however i had to use read table for the second exercise as it was whitespace delimited. then i mapped the ensembl ids the same as the worked example. i took the first column of Ensembl IDs then i used biomart to map to HGNC symbols. for 6.1 i removed the version suffix like in the worked example however it was not necessary for 6.2.



