# Exercises
These are intended to be done **after** completing the worked examples.

## Exercise 1 — https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE119732

Using **GSE119732**, 

  1. Map the distribution of the samples as a box plot and a density plot using only the raw data.


```{r ex1_get_data}
gse1 <- "GSE119732"
fetch_geo_supp(gse = gse1)

path1  <- file.path("data", gse1)
files1 <- list.files(
  path1,
  pattern = "\\.txt$|\\.txt.gz$|\\.tsv$|\\.tsv.gz$|\\.csv$|\\.csv.gz$",
  full.names = TRUE,
  recursive = TRUE
)

files1 #checking the file

x1 <- safe_read(files1[1])

head(x1)[, 1:min(6, ncol(x1))]

counts1 <- as.matrix(x1[, 2:ncol(x1)])
plot_box(counts1, main = paste(gse1, "- raw counts"))
plot_density(counts1, main = paste(gse1, "- raw counts"))
```
2.  Add colours to the plot separating the different variables in the model
  
```{r ex1_design} 
samples1 <- colnames(x1)[-1]

# parsing

patient  <- factor(substr(samples1, 1, 1))
subset   <- factor(substr(samples1, 2, 2), levels = c("1","2","3","4","5"))
sample_data1 <- data.frame(
  sample   = samples1,
  patient  = patient,
  subset   = subset
)

#sample_data1 #checking hoe i parsed it 

#design matrix
design1 <- model.matrix(~ patient + subset, data = sample_data1)
rownames(design1) <- sample_data1$sample

design1[1:10, ]

```

```{r ex1_boxplot} 
# Boxplot must be ran all at once

counts1 <- as.matrix(x1[, -1])

group_patient <- sample_data1$patient


boxplot(
  log2(counts1 + 1),
  col = as.integer(group_patient),
  las = 2,
  main = "GSE119732 raw counts by patient",
  ylab = "log2(count + 1)"
)

legend(
  "topright",
  legend = levels(group_patient),
  col    = seq_along(levels(group_patient)),
  pch    = 15,
  bty    = "n",
  title  = "patient"
)


```



```{r ex1_densityplot} 

# Density plot must be ran all at once

counts1 <- as.matrix(x1[, -1])

group_patient <- sample_data1$patient

plot(NULL,
     xlim = c(0,20),
     ylim = c(0,0.4),
     xlab = "log2(count + 1)",
     ylab = "Density",
     main = "GSE119732 density by patient")

for (i in seq_len(ncol(counts1))) {
  lines(
    density(log2(counts1[, i] + 1)),
    col = as.integer(group_patient)[i]
  )
}

legend("topright",
       legend = levels(group_patient),
       col = seq_along(levels(group_patient)),
       lty = 1,
       bty = "n",
       title = "patient")


```
  
3.  Filter out lowly expressed genes and re plot the distributions.

```{r ex1_filterandreplot}
dge1 <- DGEList(counts = counts1)

design1 <- model.matrix(~ patient + subset, data = sample_data1)

rownames(design1) <- sample_data1$sample


keep1 <- filterByExpr(dge1, design = design1)
dge1_filtered <- dge1[keep1, , keep.lib.sizes = FALSE]

cpm1_filtered <- cpm(dge1_filtered, log = FALSE)

plot_box(cpm1_filtered,
         main = paste(gse1, "- CPM after filtering (with design)"))
plot_density(cpm1_filtered,
             main = paste(gse1, "- CPM after filtering (with design)"))

```

## Exercise 2 — https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122380

Using **GSE122380**, confirm whether the ID column contains Ensembl IDs with version suffixes.

```{r ex2_get_data}
gse2 <- "GSE122380"
fetch_geo_supp(gse = gse2)

path2  <- file.path("data", gse2)
files2 <- list.files(
  path2,
  pattern = "\\.txt$|\\.txt.gz$|\\.tsv$|\\.tsv.gz$|\\.csv$|\\.csv.gz$",
  full.names = TRUE,
  recursive = TRUE
)

files2 #all good

x2 <- safe_read(files2[1])

head(x2)[, 1:min(6, ncol(x2))]

id_col_name <- colnames(x2)[1]

head_ids <- head(x2[[id_col_name]])
head_ids
all(grepl("^ENSG[0-9]+\\.[0-9]+$", head_ids))

```
as you can see there are no ensembl IDs with version suffixes

 1. Map the distribution of the samples as a box plot and a density plot using only the raw data.
 
```{r ex2_plotting}
 counts2 <- as.matrix(x2[, -1])
 plot_box(counts2, main = paste(gse2, "- raw counts"))
plot_density(counts2, main = paste(gse2, "- raw counts"))
 
```
 
 
 
 
  2.  Add colours to the plot separating the different variables in the model
  
```{r ex2_workingwiththedata} 
  
samples2 <- colnames(counts2)

parts2 <- strsplit(samples2, "_")

individual2 <- sapply(parts2, `[`, 1)
time2 <- sapply(parts2, `[`, 2)

sample_data2 <- data.frame(
  sample = samples2,
  individual = factor(individual2),
  time = factor(time2)
)

#sample_data2 #all clear 

design2 <- model.matrix(~ individual + time, data = sample_data2)
rownames(design2) <- sample_data2$sample

```


```{r ex2_boxplot} 
group_individual <- sample_data2$individual

boxplot(
  log2(counts2 + 1),
  col = as.integer(group_individual),
  las = 2,
  main = "Raw counts coloured by individual",
  ylab = "log2(count + 1)"
)

legend(
  "topright",
  legend = levels(group_individual),
  col = seq_along(levels(group_individual)),
  pch = 15,
  bty = "n",
  title = "individual"
)

```

```{r ex2_desityplot} 
plot(NULL,
     xlim = c(0,20),
     ylim = c(0,0.4),
     xlab = "log2(count + 1)",
     ylab = "Density",
     main = "Density coloured by individual")

for (i in seq_len(ncol(counts2))) {
  lines(
    density(log2(counts2[, i] + 1)),
    col = as.integer(group_individual)[i]
  )
}

legend("topright",
       legend = levels(group_individual),
       col = seq_along(levels(group_individual)),
       lty = 1,
       bty = "n",
       title = "individual")



```
  
  
  
3.  Filter out lowly expressed genes and re plot the distributions.


```{r}
dge2 <- DGEList(counts = counts2)

keep2 <- filterByExpr(dge2, design = design2)

dge2_filtered <- dge2[keep2, , keep.lib.sizes = FALSE]

cpm2_filtered <- cpm(dge2_filtered)

plot_box(cpm2_filtered)
plot_density(cpm2_filtered)

```







## Exercise 3 - 
 
Can you use the worked example to process the above two GEO records?  How?

i was following the basic premise of the worked example however for the first exercize which is based on https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE119732 . the design for this data set is 4-5 different CD4 T cell populations from 6 donors. after adjusting to the apppropriate parameters for the design i follwed the standard protocol for making boxplots, density plots, and filtering out lowly expressed genes.
 for example 2 the sample names encode two variables, the design was based on individual and time point. i separated these using strsplit() then coloured the boxplots and density plots using these variables to visualize differences between individuals and time points.
 









